Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2018-01-29 aosd13/paperBody.tex
Move contributions into parts
Already done for Part I earlier, do it for Part II.
Add better-named labels (ch:aosd13-...) to aosd13 chapters
aosd13: rephrase against overfull hboxes

\chapter{Introduction}
\label{sec:aosd13-intro}
In-memory collections of data often need efficient processing. For on-disk
data, efficient processing is already provided by database management systems~(DBMS) thanks to
their query optimizers, which support many optimizations specific to the domain
of collections. Moving in-memory data to DBMSs, however,
typically does not improve performance~\citep{Stonebraker07}, and query optimizers cannot
be reused separately since DBMS are typically monolithic and their optimizers deeply integrated.
A few collection-specific optimizations, such as shortcut
fusion~\citep{Gill93shortcut}, are supported by compilers for purely
functional languages such as Haskell. However, the implementation techniques for
those optimizations do not
generalize to many other ones, such as support for indexes. In
general, collection-specific optimizations are not supported by the general-purpose
optimizers used by typical (JIT) compilers.

@@

@@

As our running example through the chapter, we consider representing and querying a simple in-memory bibliography. A book has, in our schema, a title, a publisher and a list of authors. Each author, in turn, has a first and last name. We represent authors and books as instances of the Scala classes \code{Author} and \code{Book} shown in Fig.~\ref{fig:schema}. The class declarations list the type of each field: Titles, publishers, and first and last names are all stored in fields of type \code{String}. The list of authors is stored in a field of type \code{Seq[Author]}, that is, a sequence of authors -- something that would be more complex to model in a relational database.
The code fragment also defines a collection of books named \code{books}.
\begin{figure}[t]
\centering
\begin{lstlisting}
package schema
case class Author(firstName: String, lastName: String)
case class Book(title: String, publisher: String,
  authors: Seq[Author])

@@

\begin{figure}[t]
\centering
\begin{lstlisting}
case class BookData(title: String, authorName: String,
  coauthors: Int)

@@

\begin{figure}
\centering
\begin{lstlisting}
def titleFilterHandOpt(books: Set[Book],
                        publisher: String,
                        keyword: String) =
  for {
    book <- books
    if book.publisher == publisher && book.title.contains(keyword)
    author <- book.authors
  } yield (book.title, author.firstName + " " + author.lastName)
val res = titleFilterHandOpt(books,
  "Pearson Education", "Principles")
\end{lstlisting}
\caption{Composition of queries in Fig.~\ref{fig:query}, after inlining, query unnesting and hoisting.}
\label{fig:titleFilterho2}
\end{figure}

@@

\label{sec:solution}
The goal of {\LoS} is to let programmers write queries modularly and at a high level of abstraction and deal with optimization by a dedicated domain-specific optimizer. In our concrete example,
programmers should be able to write queries similar to the one in Fig.~\ref{fig:query}, but get the efficiency of the one in Fig.~\ref{fig:titleFilterho2}.
To allow this, {\LoS} overloads for-comprehensions and other constructs, such as string concatenation with \code{+} and field access \code{book.author}. Our overloads of these constructs reify the query as an expression tree. {\LoS} can then optimize this expression tree and execute the resulting optimized query.
Programmers explicitly trigger processing by {\LoS}, by adapting their queries as we describe in next subsection.

\section{Adapting a Query}
\label{subsec:adaptingaquery}
\begin{figure}
\centering
\begin{lstlisting}
import squopt._
import schema.squopt._

@@

val resQuery = titleFilterQuery(recordsQuery, "Principles")
val res = resQuery.optimize.eval
\end{lstlisting}
\caption{\LoS\ version of Fig.~\ref{fig:query}; \code{recordQuery} contains a reification of the query, \code{records} its result.}
\label{fig:reifiedQuery}
\end{figure}

@@

A call to \code{optimize} on a query will then take this index into account and rewrite the query to perform index lookup instead of scanning, if possible. For instance, the code in Fig.~\ref{fig:reifiedQuery} would be transparently rewritten by the optimizer to a query similar to the following:
\begin{lstlisting}
val indexedQuery =
  for {
    book <- idxByPublisher("Pearson Education")
    author <- book.authors
  } yield new BookData(book.title, author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}
Since dictionaries in Scala are functions, in the above code, dictionary lookup on \code{idxByPublisher} is represented simply as function application. The above code iterates over books having the desired publisher, instead of scanning the whole library, and performs the remaining computation from the original query. Although the index use in the listing above is notated as 
\code{idxByPublisher("Pearson Education")}, only the cached result of evaluating the index is used when
the query is executed, not the reified index definition.

@@

\chapter{Implementing SQuOpt}
\label{sec:implementation}

@@

More in detail, our compilation algorithm is as follows.
(a) We implement CSP by replacing embedded Scala values by references to the
function arguments; so for instance \code{List(1, 2, 3).map(x => x + 1)} becomes
the function \code{(s1: List[Int], s2: Int) => s1.map(x => x + s2)}.
(b) We look up the produced expression tree, together with the types of the
constants we just removed, in a cache mapping to the generated classes. If the
lookup fails we update the cache with the result of the next steps.
(c) We apply CSE on the expression.
(d) We convert the tree to code, compile it and load the generated code.

@@

\chapter{A deep EDSL for collection queries}
\label{sec:caseStudy}
In this chapter we discuss collections as a \emph{critical} case
study~\citep{flyvbjerg06five} for deep DSL embedding.

@@

\section{Implementation: Expressing the interface in Scala}
\label{sec:intfScala}
\input{aosd13/pepm13/caseStudyImpl}

@@

\chapter{Discussion}
\label{sec:discussion}
\input{aosd13/pepm13/caseStudyDiscussion}
\pg{TODO: mention discussion with Jacques Carette. Maybe mention that in the technical report only?}

\chapter{Related work}
\label{sec:relwork}
This chapter builds on prior work on language-integrated queries, query optimization, techniques for DSL embedding, and other works on code querying.

\section{Language-Integrated Queries}
Microsoft's Language-Integrated Query technology (\LINQ)~\citep{Meijer:2006:LRO:1142473.1142552,Bierman:2007:LTF:1297027.1297063} is similar to our work in that it also
reifies queries on collections to enable analysis and optimization. Such queries can be executed against a variety of backends (such as SQL databases or in-memory objects), and adding new back-ends is supported. Its implementation uses \emph{expression trees}, a compiler-supported
implicit conversion between expressions and their reification as a syntax tree. There are various major differences, though.
First, the support for expression trees is hard-coded into the compiler. This means that the techniques are not applicable in languages 
that do not explicitly support expression trees. More importantly, the way expression trees are created in \LINQ\ is generic and fixed.
For instance, it is not possible to create different tree nodes for method calls that are relevant to an analysis (such as the \code{map} method) than for method calls that are irrelevant for the analysis (such as the \code{toString} method). For this reason, expression trees in \LINQ\ 
cannot be customized to the task at hand and contain too much low-level information. It is well-known that this makes it quite hard to
implement programs operating on expression trees~\citep{Eini11Pain}.

@@

\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Like OptiQL, Jet does not apply typical database optimizations such as
indexing or filter hoisting.

@@

\chapter{Conclusions}

>>>>

\chapter{Introduction}
\label{sec:aosd13-intro}
\label{ch:aosd13-intro}
In-memory collections of data often need efficient processing. For on-disk
data, efficient processing is already provided by database management systems~(DBMS) thanks to
their query optimizers, which support many optimizations specific to the domain
of collections. Moving in-memory data to DBMSs, however,
typically does not improve performance~\citep{Stonebraker07}, and query optimizers cannot
be reused separately since DBMS are typically monolithic and their optimizers deeply integrated.
A few collection-specific optimizations, such as shortcut
fusion~\citep{Gill93shortcut}, are supported by compilers for purely
functional languages such as Haskell. However, the implementation techniques for
those optimizations do not
generalize to many other ones, such as support for indexes. In
general, collection-specific optimizations are not supported by the general-purpose
optimizers used by typical (JIT) compilers.

@@

\section{Contributions and summary}
\label{sec:aosd13-contributions}
\label{sec:navigating-aosd13}

@@

As our running example through the chapter, we consider representing and querying a simple in-memory bibliography. A book has, in our schema, a title, a publisher and a list of authors. Each author, in turn, has a first and last name. We represent authors and books as instances of the Scala classes \code{Author} and \code{Book} shown in Fig.~\ref{fig:schema}. The class declarations list the type of each field: Titles, publishers, and first and last names are all stored in fields of type \code{String}. The list of authors is stored in a field of type \code{Seq[Author]}, that is, a sequence of authors -- something that would be more complex to model in a relational database.
The code fragment also defines a collection of books named \code{books}.
\begin{figure}[htb]
\begin{lstlisting}
package schema
case class Author(firstName: String, lastName: String)
case class Book(title: String, publisher: String,
  authors: Seq[Author])

@@

\begin{figure}[htb]
\begin{lstlisting}
case class BookData(title: String, authorName: String,
  coauthors: Int)

@@

\begin{figure}[htb]
\begin{lstlisting}
def titleFilterHandOpt(books: Set[Book],
                        publisher: String,
                        keyword: String) =
  for {
    book <- books
    if book.publisher == publisher && book.title.contains(keyword)
    author <- book.authors
  } yield (book.title, author.firstName + " " + author.lastName)
val res = titleFilterHandOpt(books,
  "Pearson Education", "Principles")
\end{lstlisting}
\caption{Composition of queries in Fig.~\ref{fig:query}, after inlining, query unnesting and hoisting.}
\label{fig:titleFilterho2}
\end{figure}

@@

\label{ch:aosd13-solution}
\label{sec:solution}
The goal of {\LoS} is to let programmers write queries modularly and at a high level of abstraction and deal with optimization by a dedicated domain-specific optimizer. In our concrete example,
programmers should be able to write queries similar to the one in Fig.~\ref{fig:query}, but get the efficiency of the one in Fig.~\ref{fig:titleFilterho2}.
To allow this, {\LoS} overloads for-comprehensions and other constructs, such as string concatenation with \code{+} and field access \code{book.author}. Our overloads of these constructs reify the query as an expression tree. {\LoS} can then optimize this expression tree and execute the resulting optimized query.
Programmers explicitly trigger processing by {\LoS}, by adapting their queries as we describe in next subsection.

\section{Adapting a Query}
\label{subsec:adaptingaquery}
\begin{figure}[htb]
\centering
\begin{lstlisting}
import squopt._
import schema.squopt._

@@

val resQuery = titleFilterQuery(recordsQuery, "Principles")
val res = resQuery.optimize.eval
\end{lstlisting}
\caption{\LoS\ version of Fig.~\ref{fig:query}; \code{recordsQuery} contains a reification of the query, \code{records} its result.}
\label{fig:reifiedQuery}
\end{figure}

@@

A call to \code{optimize} on a query will then take this index into account and rewrite the query to perform index lookup instead of scanning, if possible. For instance, the code in Fig.~\ref{fig:reifiedQuery} would be transparently rewritten by the optimizer to a query similar to the following:
\begin{lstlisting}
val indexedQuery =
  for {
    book <- idxByPublisher("Pearson Education")
    author <- book.authors
  } yield new BookData(book.title,
    author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}
Since dictionaries in Scala are functions, in the above code, dictionary lookup on \code{idxByPublisher} is represented simply as function application. The above code iterates over books having the desired publisher, instead of scanning the whole library, and performs the remaining computation from the original query. Although the index use in the listing above is written as
%Beware: \linebreak simulates a normal break, hence it doesn't cause an
%underfull hbox. See https://tex.stackexchange.com/questions/39629/why-is-linebreak-failing#comment80030_39629.
\code{idxByPublisher("Pearson}\linebreak\code{Education")}, only the cached result of evaluating the index is used when
the query is executed, not the reified index definition.

@@

\chapter{Implementing SQuOpt}
\label{ch:aosd13-implementation}
\label{sec:implementation}

@@

More in detail, our compilation algorithm is as follows.
(a) We implement CSP by replacing embedded Scala values by references to the
function arguments; so for instance \code{List(1, 2, 3).map(}\linebreak\code{x => x + 1)} becomes
the function \code{(s1: List[Int], s2: Int) => s1.map(x => x + s2)}.
(b) We look up the produced expression tree, together with the types of the
constants we just removed, in a cache mapping to the generated classes. If the
lookup fails we update the cache with the result of the next steps.
(c) We apply CSE on the expression.
(d) We convert the tree to code, compile it and load the generated code.

@@

\chapter{A deep EDSL for collection queries}
\label{ch:aosd13-caseStudy}
\label{sec:caseStudy}
In this chapter we discuss collections as a \emph{critical} case
study~\citep{flyvbjerg06five} for deep DSL embedding.

@@

\section{Implementation: Expressing the interface in Scala}
\label{ch:aosd13-intfScala}
\label{sec:intfScala}
\input{aosd13/pepm13/caseStudyImpl}

@@

\chapter{Discussion}
\label{ch:aosd13-discussion}
\label{sec:discussion}
\input{aosd13/pepm13/caseStudyDiscussion}
\pg{TODO: mention discussion with Jacques Carette. Maybe mention that in the technical report only?}

\chapter{Related work}
\label{ch:aosd13-relwork}
\label{sec:relwork}
This chapter builds on prior work on language-integrated queries, query optimization, techniques for DSL embedding, and other works on code querying.

\section{Language-Integrated Queries}
Microsoft's Language-Integrated Query technology (\LINQ)~\citep{Meijer:2006:LRO:1142473.1142552,Bierman:2007:LTF:1297027.1297063} is similar to our work in that it also
reifies queries on collections to enable analysis and optimization. Such queries can be executed against a variety of backends (such as SQL databases or in-memory objects), and adding new back-ends is supported. Its implementation uses \emph{expression trees}, a compiler-supported
implicit conversion between expressions and their reification as a syntax tree. There are various major differences, though.
First, the support for expression trees is hard-coded into the compiler. This means that the techniques are not applicable in languages
that do not explicitly support expression trees. More importantly, the way expression trees are created in \LINQ\ is generic and fixed.
For instance, it is not possible to create different tree nodes for method calls that are relevant to an analysis (such as the \code{map} method) than for method calls that are irrelevant for the analysis (such as the \code{toString} method). For this reason, expression trees in \LINQ{}
cannot be customized to the task at hand and contain too much low-level information. It is well-known that this makes it quite hard to
implement programs operating on expression trees~\citep{Eini11Pain}.

@@

\citet{Ackermann12} present Jet, which also optimizes collection queries.
However, it
targets MapReduce-style computations in a distributed environment.
Like OptiQL, Jet does not apply typical database optimizations such as
indexing or filter hoisting.

@@

\chapter{Conclusions}
\label{ch:aosd13-concl}

### 2018-01-29 aosd13/pepm13/caseStudyImpl.tex
Fix/Cleanup setting of TeX-master
It didn't work for some files in aosd13.
aosd13: rephrase against overfull hboxes

\smartParagraph{Method calls and operators}
In our example, the operator \code{+} should be available on \code{Exp[String]}, but not on \code{Exp[Boolean]}, because \code{+} is available on \code{String} but not on \code{Boolean}.
Furthermore, we want \code{str1 + str2} to have type \code{Exp[String]} and to evaluate not to a string concatenation but to a call of \code{StringConcat}, that is, to an expression tree which \emph{represents} \code{str1 + str2}. This is a somewhat unusual requirement, because usually, the interface of a generic type does not depend on the type parameters.

@@

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
The implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

\smartParagraph{Literal values}
However, a string concatenation might also include constants, as in \code{str1 + "foo"} or \code{"bar" + str1}.
To lift \code{str1 + "foo"}, we introduce a lifting for constants which wraps them in a \code{Const} node:

@@

The compiler will now rewrite \code{str1 + "foo"} to \code{expToStringOps(str1) + pure("foo")}, and
\code{str1 + "foo" + str2} to \code{expToStringOps(str1) + pure("foo") + str2}.
Different implicit conversions cooperate successfully to lift the expression.
%Since \code{Exp[String].+} requires an argument of type \code{Exp[String]} but \code{"foo"} has type \code{String},
%the compiler wraps \code{"foo"} with a call to \code{pure}. Hence

@@

Implicit conversions for user-defined types cooperate with other ones; for instance, \code{author.firstName + " " + author.lastName} is rewritten to
\begin{lstlisting}
(expToStringOps(expToAuthorOps(author).firstName) + pure(" ")) +
  expToAuthorOps(author).lastName
\end{lstlisting}

@@

Note that reified functions have type \code{Exp[A] => Exp[B]} instead of the more regular \code{Exp[A => B]}. We chose the former over the latter to support Scala's syntax for anonymous functions and for-comprehensions which is hard-coded to produce or consume instances of the pre-defined \code{A => B} type. We have to reflect this irregularity in the lifting of methods and functions by treating the types of higher-order arguments accordingly.

@@

\begin{lstlisting}
def map[U](f: T => U)(implicit cbf: CanBuildFrom[Repr, U, That]): That
def flatMap[U](f: T => Traversable[U])(implicit cbf: CanBuildFrom[Repr, U, That]): That
def withFilter[U](p: T => Boolean): Repr.
\end{lstlisting}

@@

\smartParagraph{The lifting}
The basic idea is to use the enrich-my-library pattern to lift collection methods from \code{TraversableLike[T, Repr]} to \code{TraversableLikeOps[T, Repr]}:
\begin{lstlisting}
implicit def expToTraversableLikeOps[T, Repr](v: Exp[TraversableLike[T, Repr]]) =
  new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

Subclasses of \code{TraversableLike[T, Repr]} also subclass both \code{Traversable[T]} and \code{Repr}; to take advantage of this during interpretation and optimization, we need to restrict the type of \code{expToTraversableLikeOps}, getting the following conversion:\footnote{Due to type inference bugs, the actual implicit conversion needs to be slightly more complex, to mention \code{T} directly in the argument type. We reported the bug at \url{https://issues.scala-lang.org/browse/SI-5298}.}

%We use enrich-my-library again with an implicit conversion to \code{TraversableLikeOps[T, Repr]}, shown later, which corresponds to
%\code{TraversableLike[T, Repr]}.
%We provide an implicit conversion to \code{TraversableLikeOps[T, Repr]} from \code{Exp[Repr]} for any type \code{Repr} which is a subtype of \code{TraversableLike[T, Repr]} and \code{Traversable[T]} (for some element type \code{T}). This can be expressed as:
\begin{lstlisting}
implicit def expToTraversableLikeOps
  [T, Repr <: Traversable[T] with TraversableLike[T, Repr]](v: Exp[Repr]) =
    new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

  def map[U, That <: TraversableLike[U, That]](f: Exp[T] => Exp[U])
         (implicit c: CanBuildFrom[Repr, U, That]): Exp[That] = MapNode(this.t, FuncExp(f))
  //... other methods ...
}
//definitions of MapNode, Filter omitted.
\end{lstlisting}
\end{center}
\caption{Lifting \code{TraversableLike}}
\label{fig:liftTrav}
\end{figure}

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\section*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] => Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.

@@

%\begin{lstlisting}
%def toFOAS[S, T](funBody: Exp[S] => Exp[T]) = 
%\end{lstlisting}
In our example, \code{Fun(str => str + "!")} produces\\
\code{FOASFun(StringConcat(TypedVar[String](1), Const("!")), TypedVar[String](1))}%
%: since an unbound variable appears here, this can be easily recognized as an open term
.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../../main"
%%% End:

>>>>

\smartParagraph{Method calls and operators}
In our example, operator \code{+} should be available on \code{Exp[String]}, but not on \code{Exp[Boolean]}, because \code{+} is available on \code{String} but not on \code{Boolean}.
Furthermore, we want \code{str1 + str2} to have type \code{Exp[String]} and to evaluate not to a string concatenation but to a call of \code{StringConcat}, that is, to an expression tree which \emph{represents} \code{str1 + str2}. This is a somewhat unusual requirement, because usually, the interface of a generic type does not depend on the type parameters.

@@

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(}\linebreak\code{str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
The implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

\smartParagraph{Literal values}
However, string concatenation might also be applied to constants, as in \code{str1 + "foo"} or \code{"bar" + str1}.
To lift \code{str1 + "foo"}, we introduce a lifting for constants which wraps them in a \code{Const} node:

@@

The compiler will now rewrite \code{str1 + "foo"} to \code{expToStringOps(str1) + pure("foo")}, and
\code{str1}\linebreak\code{+ "foo" + str2} to \code{expToStringOps(str1) + pure("foo") + str2}.
Different implicit conversions cooperate successfully to lift the expression.
%Since \code{Exp[String].+} requires an argument of type \code{Exp[String]} but \code{"foo"} has type \code{String},
%the compiler wraps \code{"foo"} with a call to \code{pure}. Hence

@@

Implicit conversions for user-defined types cooperate with other ones; for instance, \code{author.}\linebreak\code{firstName + " " + author.lastName} is rewritten to
\begin{lstlisting}
(expToStringOps(expToAuthorOps(author).firstName) + pure(" ")) +
  expToAuthorOps(author).lastName
\end{lstlisting}

@@

Reified functions have type \code{Exp[A] => Exp[B]} instead of the more regular \code{Exp[A => B]}. We chose the former over the latter to support Scala's syntax for anonymous functions and for-comprehensions which is hard-coded to produce or consume instances of the pre-defined \code{A => B} type. We have to reflect this irregularity in the lifting of methods and functions by treating the types of higher-order arguments accordingly.

@@

\begin{lstlisting}
def map[U](f: T => U)(implicit cbf: CanBuildFrom[Repr, U, That]): That
def flatMap[U](f: T => Traversable[U])
  (implicit cbf: CanBuildFrom[Repr, U, That]): That
def withFilter[U](p: T => Boolean): Repr.
\end{lstlisting}

@@

\smartParagraph{The lifting}
The basic idea is to use the enrich-my-library pattern to lift collection methods from \code{TraversableLike[T, Repr]} to \code{TraversableLikeOps[T, Repr]}:
\begin{lstlisting}
implicit def expToTraversableLikeOps[T, Repr]
    (v: Exp[TraversableLike[T, Repr]]) =
  new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

Subclasses of \code{TraversableLike[T, Repr]} also subclass both \code{Traversable[T]} and \code{Repr}; to take advantage of this subclass relation during interpretation and optimization, we need to restrict the type of \code{expToTraversableLikeOps}, obtaining the following conversion:\footnote{Due to type inference bugs, the actual implicit conversion needs to be slightly more complex, to mention \code{T} directly in the argument type. We reported the bug at \url{https://issues.scala-lang.org/browse/SI-5298}.}

%We use enrich-my-library again with an implicit conversion to \code{TraversableLikeOps[T, Repr]}, shown later, which corresponds to
%\code{TraversableLike[T, Repr]}.
%We provide an implicit conversion to \code{TraversableLikeOps[T, Repr]} from \code{Exp[Repr]} for any type \code{Repr} which is a subtype of \code{TraversableLike[T, Repr]} and \code{Traversable[T]} (for some element type \code{T}). This can be expressed as:
\begin{lstlisting}
implicit def expToTraversableLikeOps[T,
    Repr <: Traversable[T] with TraversableLike[T, Repr]](v: Exp[Repr]) =
  new TraversableLikeOps[T, Repr](v)
\end{lstlisting}

@@

  def map[U, That <: TraversableLike[U, That]](f: Exp[T] => Exp[U])
         (implicit c: CanBuildFrom[Repr, U, That]): Exp[That] =
    Node(this.t, FuncExp(f))
  //... other methods ...
}
//definitions of MapNode, Filter omitted.
\end{lstlisting}
\end{center}
\caption{Lifting \code{TraversableLike}}
\label{fig:liftTrav}
\end{figure}

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\section*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] =>}\linebreak\code{Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.

@@

%\begin{lstlisting}
%def toFOAS[S, T](funBody: Exp[S] => Exp[T]) =
%\end{lstlisting}
In our example, \code{Fun(str => str + "!")} produces
\begin{lstlisting}
FOASFun(StringConcat(TypedVar[String](1), Const("!")),
  TypedVar[String](1))
\end{lstlisting}
%: since an unbound variable appears here, this can be easily recognized as an open term

@@
